---
title: Applied Statistical Methods - `r tools::toTitleCase(params$doctype)` 4
author: Peter von Rohr
date: 2022-03-16
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
    keep_md: false
    keep_tex: false
header-includes:
 \usepackage{longtable}
 \usepackage{float}
params:
  doctype:
    label: 'Document Type'
    value: 'solution'
    choices: ['exercise', 'solution', 'notebook']
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
s_ex04p01_data_path <- "https://charlotte-ngs.github.io/asmss2022/data/asm_bw_flem.csv"
```


## Problem 1: Overfitting
Use the extended dataset on `Body Weight` of animals and fit all the variables and the factor breed. Compare the result with a regression that uses only `Breast Circumference` or with the linear model that only uses the factor `Breed`. The data set is available from: `r s_ex04p01_data_path` 



<!-- your-solution-start
### Your Solution

* Read the data

* Fit the full model

* Fit the model with only `Breast Circumference`

* Fit the model with only `Breed`


---  your-solution-end -->


<!-- master-solution-start -->

### Solution

* Read the data

```{r}
s_ex04p01_data_path <- "https://charlotte-ngs.github.io/asmss2022/data/asm_bw_flem.csv"
tbl_ex04p01_data <- readr::read_csv(file = s_ex04p01_data_path)
```

* Fit the full model

```{r}
lm_ex04p01_full <- lm(formula = `Body Weight` ~ `Breast Circumference` + BCS + HEI + Breed, data = tbl_ex04p01_data)
summary(lm_ex04p01_full)
```

* Fit the model with only `Breast Circumference`

```{r}
lm_ex04p01_bwbc <- lm(formula = `Body Weight` ~ `Breast Circumference`, data = tbl_ex04p01_data)
summary(lm_ex04p01_bwbc)
```

* Fit only the model with the factor `Breed`

```{r}
lm_ex04p01_bwbreed <- lm(formula = `Body Weight` ~ Breed, data = tbl_ex04p01_data)
summary(lm_ex04p01_bwbreed)
```

<!-- master-solution-end -->


## Problem 2

<!-- your-solution-start
### Your Solution

---  your-solution-end -->


<!-- master-solution-start -->

### Solution

<!-- master-solution-end -->
 
