---
title: Applied Statistical Methods - `r tools::toTitleCase(params$doctype)` 4
author: Peter von Rohr
date: 2022-03-16
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
    keep_md: false
    keep_tex: false
header-includes:
 \usepackage{longtable}
 \usepackage{float}
params:
  doctype:
    label: 'Document Type'
    value: 'solution'
    choices: ['exercise', 'solution', 'notebook']
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
s_ex04p01_data_path <- "https://charlotte-ngs.github.io/asmss2022/data/asm_bw_flem.csv"
```


## Problem 1: Overfitting
Use the extended dataset on `Body Weight` of animals and fit all the variables and the factor breed. Compare the result with a regression that uses only `Breast Circumference` or with the linear model that only uses the factor `Breed`. The data set is available from: `r s_ex04p01_data_path` 



<!-- your-solution-start
### Your Solution

* Read the data

* Fit the full model

* Fit the model with only `Breast Circumference`

* Fit the model with only `Breed`


---  your-solution-end -->


<!-- master-solution-start -->

### Solution

* Read the data

```{r}
s_ex04p01_data_path <- "https://charlotte-ngs.github.io/asmss2022/data/asm_bw_flem.csv"
tbl_ex04p01_data <- readr::read_csv(file = s_ex04p01_data_path)
```

* Fit the full model

```{r}
lm_ex04p01_full <- lm(formula = `Body Weight` ~ `Breast Circumference` + BCS + HEI + Breed, data = tbl_ex04p01_data)
summary(lm_ex04p01_full)
```

* Fit the model with only `Breast Circumference`

```{r}
lm_ex04p01_bwbc <- lm(formula = `Body Weight` ~ `Breast Circumference`, data = tbl_ex04p01_data)
summary(lm_ex04p01_bwbc)
```

* Fit only the model with the factor `Breed`

```{r}
lm_ex04p01_bwbreed <- lm(formula = `Body Weight` ~ Breed, data = tbl_ex04p01_data)
summary(lm_ex04p01_bwbreed)
```

The comparison of the models shows that the full model does not produce a better model fit. The reason for this is that the explanatory variables in the full model are correlated among each other. As a result of this correlation structure, the same information is contained in different variables and as a result the single variables do not contribute a substantial amount to the explanation of the variation in the response variable. 

The correlation structure amoung the different variables can be visualized via a so called `pairs` plot. 

```{r}
tbl_ex04p01_data$Breed <- as.factor(tbl_ex04p01_data$Breed)
pairs(formula = ~ `Breast Circumference` + BCS + HEI + Breed , data = tbl_ex04p01_data)
```

From this plot, we can clearly see that `Breast Circumference` and `Breed` are correlated. If we switch levels 2 and 3 of the breeds, then we can see the relationship between `Breast Circumference` and `Breed` even better.

<!-- master-solution-end -->


## Problem 2: Plotting
The first step before doing any analysis should always be to plot the data which helps to visualise the internal structure of a dataset. A very instructive plot is the so-called `pairs`-plot. This plot can be done using the function `pairs()`. The task of this problem is to create a `pairs`-plot for the extended dataset on `Body Weight` of animals. The input to the function `pairs()` must be all numeric. This means that the column containing the `Breed` in our dataset must be converted to a datatype called `factor`. This can be done using the function `as.factor()`. 

Results of linear models can also be plotted. In such plots, we are mainly interested in the behavior of the residuals. Hence, fit a linear regression model between `Body Weight` and `Breast Circumference` and plot the resulting linear model object.

<!-- your-solution-start
### Your Solution

* Read the dataset

* Convert the breed column to a factor

* Create a `pairs`-plot

* Fit the linear regression model

* Plot the result

---  your-solution-end -->


<!-- master-solution-start -->

### Solution

* Read the dataset

```{r}
s_ex04p02_data_path <- "https://charlotte-ngs.github.io/asmss2022/data/asm_bw_flem.csv"
tbl_ex04p02_data <- readr::read_csv(file = s_ex04p02_data_path)
```


* Convert the breed column to a factor

```{r}
tbl_ex04p02_data$Breed <- as.factor(tbl_ex04p02_data$Breed)
```

* Create a `pairs`-plot

```{r}
pairs(tbl_ex04p02_data)
```

The above matrix of scatterplots shows relationships between pairs of variables. 

* Fit the linear regression model

```{r}
lm_ex04p02 <- lm(formula = `Body Weight` ~ `Breast Circumference`, data = tbl_ex04p02_data)
```


* Plot the result

```{r}
plot(lm_ex04p02)
```

For the behavior of the resiudals, we are focusing on the first two plots. The first plot shows whether there is a dependence pattern between the residuals and the fitted values. For this plot a random pattern is desired. The second plot shows a QQ-plot of the residuals. This plot shows any deviation of the numeric distribution of the residuals from the normal distribution.

<!-- master-solution-end -->
 
