---
title: Applied Statistical Methods - `r tools::toTitleCase(params$doctype)` 5
author: Peter von Rohr
date: 2022-03-26
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
    keep_md: false
    keep_tex: false
header-includes:
 \usepackage{longtable}
 \usepackage{float}
params:
  doctype:
    label: 'Document Type'
    value: 'solution'
    choices: ['exercise', 'solution', 'notebook']
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, echo=FALSE}
s_ex05p01_data_path <- "https://charlotte-ngs.github.io/asmss2022/data/asm_bw_flem.csv"
```

## Problem 1: Helmert Contrasts
Use the dataset of `Body Weight` and `Breed` to fit a linear model. The aim of this exercise is to use the `Helmert`-contrasts instead of the defautl `Treatment` contrasts. What are the estimable functions used in the `Helmert`-Contrasts and what are the effects that are reported for the different levels of the factor `Breed`? Verify your answer by comparing estimable functions of solutions of the least squares normal equations to the effects of `lm()`.

The dataset is available under

```{r, echo=FALSE}
cat(s_ex05p01_data_path, "\n")
```

### Hint

* Use `options(contrasts = c("contr.helmert", "contr.helmert"))` to change the default contrasts to the desired `Helmert`-Contrasts


<!-- your-solution-start
### Your Solution

* Compute solutions to least squares normal equation

* Change contrasts from default to Helmert

* Find estimable functions associated to Helmert contrasts

* Check back with effects of `lm()`

---  your-solution-end -->


<!-- master-solution-start -->

### Solution

* Compute solutions to least squares normal equation

* Change contrasts from default to Helmert

* Find estimable functions associated to Helmert contrasts

* Check back with effects of `lm()`

<!-- master-solution-end -->


## Problem 2

<!-- your-solution-start
### Your Solution

---  your-solution-end -->


<!-- master-solution-start -->

### Solution

<!-- master-solution-end -->


## Problem 3

<!-- your-solution-start
### Your Solution

---  your-solution-end -->


<!-- master-solution-start -->

### Solution

<!-- master-solution-end -->
 
