---
title: Applied Statistical Methods - Notebook 5
author: Peter von Rohr
date: '2022-03-26'
output: html_notebook
params:
  doctype:
    label: Document Type
    value: solution
    choices:
    - exercise
    - solution
    - notebook
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, echo=FALSE}
s_ex05p01_data_path <- "https://charlotte-ngs.github.io/asmss2022/data/asm_bw_flem.csv"
# s_ex05p01_data_path <- file.path(here::here(), "docs", "data", "asm_bw_flem.csv")
```

## Problem 1: Helmert Contrasts
Use the dataset of `Body Weight` and `Breed` to fit a linear model of `Body Weight` on `Breed`. The aim of this exercise is to use the `Helmert`-contrasts instead of the defautl `Treatment` contrasts. What are the estimable functions used in the `Helmert`-Contrasts and what are the effects that are reported for the different levels of the factor `Breed`? Verify your answer by comparing estimable functions of solutions of the least squares normal equations to the effects of `lm()`.

The dataset is available under

```{r, echo=FALSE}
cat(s_ex05p01_data_path, "\n")
```

### Hint

* Use `options(contrasts = c("contr.helmert", "contr.helmert"))` to change the default contrasts to the desired `Helmert`-Contrasts


### Your Solution

* Compute solutions to least squares normal equation

* Change contrasts from default to Helmert

* Find estimable functions associated to Helmert contrasts

* Check back with effects of `lm()`





```{r, echo=FALSE}
vec_nobs <- c(10,30,100)
s_ex05p02_data_path <- "https://charlotte-ngs.github.io/asmss2022/data/asm_bw_flem.csv"
# s_ex05p02_data_path <- file.path(here::here(), "docs", "data", "asm_bw_flem.csv")
```

## Problem 2: Simulation
Use the results of the regression of `Body Weight` on `Breast Circumference` and simulate three datasets with `r vec_nobs[1]`, `r vec_nobs[2]` and `r vec_nobs[3]` observations respectively. What is the number of observations required to obtain the same regression results from the simulated dataset that you used in the simulation? 

The original dataset is available under:

```{r, echo=FALSE}
cat(s_ex05p02_data_path, "\n")
```

### Your Solution

* Run the regression analysis of `Body Weight` on `Breast Circumference`

* Create the three datasets

* Analyse each of the datasets using `lm()`

* Check the obtained results, whether they match the input from the simulation







```{r, echo=FALSE, results='asis'}
cat('\n---\n\n _Latest Changes: ', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), ' (', Sys.info()['user'], ')_\n', sep = '')
```
 
