---
title: Applied Statistical Methods - `r tools::toTitleCase(params$doctype)` 5
author: Peter von Rohr
date: 2022-03-26
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
    keep_md: false
    keep_tex: false
header-includes:
 \usepackage{longtable}
 \usepackage{float}
params:
  doctype:
    label: 'Document Type'
    value: 'solution'
    choices: ['exercise', 'solution', 'notebook']
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, echo=FALSE}
s_ex05p01_data_path <- "https://charlotte-ngs.github.io/asmss2022/data/asm_bw_flem.csv"
# s_ex05p01_data_path <- file.path(here::here(), "docs", "data", "asm_bw_flem.csv")
```

## Problem 1: Helmert Contrasts
Use the dataset of `Body Weight` and `Breed` to fit a linear model of `Body Weight` on `Breed`. The aim of this exercise is to use the `Helmert`-contrasts instead of the defautl `Treatment` contrasts. What are the estimable functions used in the `Helmert`-Contrasts and what are the effects that are reported for the different levels of the factor `Breed`? Verify your answer by comparing estimable functions of solutions of the least squares normal equations to the effects of `lm()`.

The dataset is available under

```{r, echo=FALSE}
cat(s_ex05p01_data_path, "\n")
```

### Hint

* Use `options(contrasts = c("contr.helmert", "contr.helmert"))` to change the default contrasts to the desired `Helmert`-Contrasts


<!-- your-solution-start
### Your Solution

* Compute solutions to least squares normal equation

* Change contrasts from default to Helmert

* Find estimable functions associated to Helmert contrasts

* Check back with effects of `lm()`

---  your-solution-end -->


<!-- master-solution-start -->

### Solution

* Compute solutions to least squares normal equation. First the data is read from the given file

```{r}
tbl_e05p01 <- readr::read_csv(file = s_ex05p01_data_path)
```

The data is sorted according to the breed

```{r}
tbl_e05p01 <- tbl_e05p01[order(tbl_e05p01$Breed),]
tbl_e05p01
```

A solution vector depends on the matrix $X$ and on the vector $y$. The vector $y$ is directly obtained from the column `Body Weight` of the dataframe. 

```{r}
vec_y <- tbl_e05p01$`Body Weight`
```

The matrix $X$ can be obtained from the function `model.matrix()`. 

```{r}
mat_X <- model.matrix(lm(`Body Weight` ~ 0 + Breed, data = tbl_e05p01))
mat_X <- cbind(matrix(1, nrow = nrow(mat_X), ncol = 1), mat_X)
mat_X
```

A solution for the least squares normal equations is obtained by 

```{r}
mat_xtx_ginv <- MASS::ginv(crossprod(mat_X))
mat_xty <- crossprod(mat_X, vec_y)
mat_b0 <- crossprod(mat_xtx_ginv, mat_xty)
mat_b0
```

The solutions correspond to the vector $b^0$ with the components

```{r, echo=FALSE, results='asis'}
vec_b0 <- c("\\mu", "\\alpha_1", "\\alpha_2", "\\alpha_3")
cat("$$\n")
cat("b^0 = ")
cat(paste0(rmdhelp::bcolumn_vector(pvec = vec_b0), collapse = "\n"), "\n")
cat(" = \n")
cat(paste0(rmdhelp::bmatrix(pmat = round(mat_b0, digits = 3)), collapse = "\n"), "\n")
cat("$$\n")
```


* Change contrasts from default to Helmert. Start by saving away the existing options

```{r}
opts <- options()
```

Change contrasts

```{r}
options(contrasts = c("contr.helmert", "contr.helmert"))
getOption("contrasts")
```


* Find estimable functions associated to Helmert contrasts. Estimable functions are derived from the contrasts matrix. First the Breed column must be converted to a factor.

```{r}
tbl_e05p01$Breed <- as.factor(tbl_e05p01$Breed)
c_mat_helmert <- contrasts(tbl_e05p01$Breed)
c_mat_helmert
```

Add a columns of all ones to `c_mat_helmert`. 

```{r}
c_mat_helmert <- cbind(matrix(1, nrow = nrow(c_mat_helmert), ncol = 1), c_mat_helmert)
c_mat_helmert
```

Compute the inverse of `c_mat_helmert`

```{r}
est_mat_helmert <- solve(c_mat_helmert)
est_mat_helmert
```

The first row tells us how the intercept is computed. The intercept ($\hat{b}_0$) here corresponds to 

$$\hat{b}_0 = {1\over 3}\left( E(y_{1.}) + E(y_{2.}) + E(y_{3.})\right)$$
where $E(y_{1.})$,  $E(y_{2.})$ and $E(y_{3.})$ are the mean values of `Body Weight` for Angus, Limousin and Simmental animals, respectively. 

```{r}
n_mean_angus <- mean(tbl_e05p01[tbl_e05p01$Breed == "Angus", ]$`Body Weight`)
n_mean_limousin <- mean(tbl_e05p01[tbl_e05p01$Breed == "Limousin", ]$`Body Weight`)
n_mean_simmental <- mean(tbl_e05p01[tbl_e05p01$Breed == "Simmental", ]$`Body Weight`)
mean(c(n_mean_angus, n_mean_limousin,n_mean_simmental))
```

The second row of `est_mat_helmert` shows the first estimable function that is used. It corresponds to 

$$\hat{b}_1 = {1\over 2}(\alpha_2 - \alpha_1)$$
where $\hat{b}_1$ measures the difference between the breeds `Limousin` and `Angus` corresponding to

```{r}
1/2*(mat_b0[3] - mat_b0[2])
```

The third row of `est_mat_helmert` shows how the `Body Weight` of the breed `Simmental` is compared to the two other breeds. It is 

$$\hat{b}_2 = {1\over 6}(2\alpha_3 - \alpha_2 - \alpha_1)$$

which measures the difference between `Simmental` and `Limousin` and `Angus` together.

```{r}
1/6 * (2*mat_b0[4] - mat_b0[3] - mat_b0[2])
```


* Check back with effects of `lm()`. The estimate for the intercept is

```{r}
lm_helmert <- lm(`Body Weight` ~ Breed, data = tbl_e05p01)
coefficients(lm_helmert)["(Intercept)"]
```

The estimates for the `Breed` effects can be seen from the list of all coefficients.

```{r}
coefficients(lm_helmert)
```



* Restore original options


```{r}
options(opts)
```

<!-- master-solution-end -->


```{r, echo=FALSE}
vec_nobs <- c(10,30,100)
s_ex05p02_data_path <- "https://charlotte-ngs.github.io/asmss2022/data/asm_bw_flem.csv"
# s_ex05p02_data_path <- file.path(here::here(), "docs", "data", "asm_bw_flem.csv")
```

## Problem 2: Simulation
Use the results of the regression of `Body Weight` on `Breast Circumference` and simulate three datasets with `r vec_nobs[1]`, `r vec_nobs[2]` and `r vec_nobs[3]` observations respectively. What is the number of observations required to obtain the same regression results from the simulated dataset that you used in the simulation? 

The original dataset is available under:

```{r, echo=FALSE}
cat(s_ex05p02_data_path, "\n")
```

<!-- your-solution-start
### Your Solution

* Run the regression analysis of `Body Weight` on `Breast Circumference`

* Create the three datasets

* Analyse each of the datasets using `lm()`

* Check the obtained results, whether they match the input from the simulation

---  your-solution-end -->


<!-- master-solution-start -->

### Solution

* Run the regression analysis of `Body Weight` on `Breast Circumference`

```{r}
tbl_bwbc <- readr::read_csv(file = s_ex05p02_data_path)
lm_bwbc <- lm(`Body Weight` ~ `Breast Circumference`, data = tbl_bwbc)
b0 <- coefficients(lm_bwbc)["(Intercept)"]
b1 <- coefficients(lm_bwbc)["`Breast Circumference`"]
mean_bc <- mean(tbl_bwbc$`Breast Circumference`)
sd_bc <- sd(tbl_bwbc$`Breast Circumference`)
sry_bwbc <- summary(lm_bwbc)
sd_res <- sry_bwbc$sigma
```


* Create the three datasets. In a first step, we create a function that takes as arguments 
    + number of observations
    + intercept $b_0$
    + slope $b_1$
    + mean `Breast Circumference`
    + standard deviation `Breast Circumference`
    + standard deviation of residuals

and returns a dataset according to these input values.

```{r}
simulate_bwbc <- function(pn_nrobs, pn_b0, pn_b1, pn_mean_x, pn_sd_x, pn_sd_res){
  vec_bc <- rnorm(pn_nrobs, mean = pn_mean_x, sd = pn_sd_x)
  vec_bw <- pn_b0 + pn_b1 * vec_bc + rnorm(pn_nrobs, mean = 0, sd = pn_sd_res)
  tbl_result <- tibble::tibble(Animal = c(1:pn_nrobs),
                               BC = vec_bc,
                               BW = vec_bw)
  return(tbl_result)
}
```

With the above defined function, we can create a list with the three datasets

```{r}
set.seed(1928)
vec_nobs <- c(10,30,100)
l_data_set <- lapply(vec_nobs, simulate_bwbc, 
                     pn_b0 = b0, 
                     pn_b1 = b1, 
                     pn_mean_x = mean_bc, 
                     pn_sd_x = sd_bc, 
                     pn_sd_res = sd_res) 
```

Each of the datasets is analysed by `lm()` and the results are again stored in a list

```{r}
l_lm_result <- lapply(l_data_set, function(x) lm(BW ~ BC, data = x))
```

Collect the resutls into a table

```{r}
tbl_result <- NULL
for (cur_res in l_lm_result){
  sry_cur_res <- summary(cur_res)
  tbl_cur <- tibble::tibble(NrObs = length(cur_res$residuals),
              Intercept_Estimate = coefficients(sry_cur_res)["(Intercept)", "Estimate"],
              Intercept_StdErr = coefficients(sry_cur_res)["(Intercept)", "Std. Error"],
              Slope_Estimate = coefficients(sry_cur_res)["BC", "Estimate"],
              Slope_StdErr = coefficients(sry_cur_res)["BC", "Std. Error"],
              ResStdErr = sry_cur_res$sigma)
  if (is.null(tbl_result)){
    tbl_result <- tbl_cur
  } else {
    tbl_result <- dplyr::bind_rows(tbl_result, tbl_cur)
  }
}
knitr::kable(tbl_result)
```

The true values used in the sumulation are 

```{r, echo=FALSE, results='asis'}
tbl_true_values <- tibble::tibble(Intercept = b0,
                                  Slope = b1,
                                  ResStdErr = sd_res)
knitr::kable(tbl_true_values,
             longtable = TRUE,
             booktabs = TRUE)
```


<!-- master-solution-end -->


