---
title: Contrasts, Plots and Simulation
author: Peter von Rohr
date: "`r Sys.Date()`"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Three Topics

1. Contrasts in R
2. Plots
3. Simulation


## Contrasts in R

* Estimable functions used in R are encoded by so-called `contrasts`
* Use `getOption("contrasts")` to find which contrasts are used
* Use `options(contrasts = c("<contast_unordered>", "<contrast_ordered"))` to change contrasts


## Contrasts and Estimable Functions

```{r, echo=FALSE, message=FALSE, warning=FALSE}
s_bwbr_path <- "https://charlotte-ngs.github.io/asmss2022/data/asm_bw_flem.csv"
s_bwbr_path <- file.path(here::here(), "docs", "data", "asm_bw_flem.csv")
tbl_bwbr <- readr::read_csv(file = s_bwbr_path)
```

* Relationship between contrasts and estimable functions via __contrasts matrix__

```{r}
(cm_treat <- contrasts(as.factor(tbl_bwbr$Breed)))
```


## Estimable Functions

* Extend contrast matrix

```{r}
(cm_treat <- cbind(matrix(1, 
                     nrow = nrow(cm_treat),
                     ncol = 1), 
                   cm_treat))
```


## Estimable Functions II

* Invert extended contrast matrix

```{r}
(em_treat <- solve(cm_treat))
```


## Intercept

```{r, echo=FALSE}
n_mean_angus <- mean(tbl_ex03p03_data[tbl_ex03p03_data$Breed == "Angus", ]$`Body Weight`)
n_mean_limousin <- mean(tbl_ex03p03_data[tbl_ex03p03_data$Breed == "Limousin", ]$`Body Weight`)
n_mean_simmental <- mean(tbl_ex03p03_data[tbl_ex03p03_data$Breed == "Simmental", ]$`Body Weight`)

m <- 
```

* First row of `em_treat` shows computation of estimate for intercept
* Define vector $m$ as the vector of the mean values for `Body Weight` for all breeds




