---
title: "Master Thesis Topics with Qualitas AG and ETHZ"
author: "Peter von Rohr"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<style type="text/css">
h1.title {
  font-size: 32px;
  color: Black;
}
h1 { /* Header 1 */
  font-size: 28px;
  color: DarkRed;
}
h2 { /* Header 2 */
    font-size: 22px;
  color: DarkRed;
}
h3 { /* Header 3 */
  font-size: 18px;
  color: DarkRed;
}
</style>

If you are interested in any of the following topics or if you have questions, please contact Peter von Rohr (Qualitas AG) Chamerstrasse 56, 6300 Zug; peter.vonrohr@qualitasag.ch, Tel 041 768 92 21.



### New Breeding Programs {#project_header .emphasized}
Swiss Dairy Cattle breeding organisations have traditionally been collecting the same type of data from a large number of farms. Recently the breeding organisations have diversified their services and now offer different type of memberships. Depending on the type of membership that a farmer is committing to different types of data are collected. Some of these newly collected data offer the possibility to use the information from this data as selection tools. To the best of our knowledge there has not been any scientific investigation about the benefits of collecting the new types of data for the breeding program as a whole.


### Connectedness Between Herds In Different Populations {#project_header .emphasized}
In populations with a low rate of artificial insemination, it is often a problem that effects of herds cannot be compared or estimated. There are different measures that indicate such problems. These measures are all contained in the general concept of connectedness.


### Resilience Traits Based on Milk Yields From Robotic Milking Systems {#project_header .emphasized}
More than 100 farms with robotic milking systems are connected to the service of automated data exchange. For those farms, the milk yield per cow is collected on a daily basis. This allows to describe the trajectory of the trait milk yield over time with much more accuracy. This new type of automated data collection is expected to deliver a lot more information about different traits of a cow that are related to health and resilience. This project aims at discovering such new traits which can ultimately be used as new traits for selection in dairy populations. 


### Evaluation of Fertalys Data
Fertalys is a pregnancy testing service offered by the Swiss Dairy Cattle Breeding Organisations together with Swisslab. Results from Fertalys tests might contain information about different relevant fertility traits. The goals of this project is to find possible connections between test results and other traits.


### First steps towards a genetic evaluation for conception rate
An analysis comparable to https://doi.org/10.3168/jds.2018-15858 is to be developed.


### Early Maturity in Swiss Beef Cattle
Early maturity is an important trait in beef cattle. But this trait is not easy to integrate into a breeding program. Different solutions to this problem should be proposed and evaluated. An alternative to the currently considered auxillary trait 'fat coverate' is the approach based on 'DAGE'. 


### Application of Generalized Linear Mixed Effect Models
Based on the results of a previous master thesis, different comparison studies should be undertaken where conventional linear mixed models are compared with the generalized linear mixed model based on the same data set. 


### Relationship between Myostatin-Status with Calving Traits
Some beef cattle breeds (Charolais, Limousin, ...) are looking into the genetic basis of the myostatin-gene (double-muscle). Recent developments in genomic selection of beef cattle in Switzerland give the possibility to analyse the status of the myostatin-gene in the Swiss beef cattle breeds. A specific emphasis is to be put to the relationship with other important characteristics of the animals such as calving traits.


### Early first calving evaluation
This trait is associated with efficiency and with environmental footprint of animals. A number of publications already exist on this topic. Tasks would consist of 

* descriptive statistics in Swiss cattle populations
* estimation of genetic parameters
* characterisation of different populations at the genomic level (GWAS, ...)


### PSB-Measures
The Swiss Federal government has decided on measures to reduce the amount of phosphoros (P) and nitrogen (N) that is released to the environment due to livestock production. The possible impact of such measures is to be investigated. 


### Feed Efficiency
In the more recent past huge efforts are undertaken to improve feed efficiency of dairy cows. Thies is due to the fact that the economic pressure for dairy farms have dramatically increased during the past few years and also because of the higher awareness of the impact of livestock production on the environment. In this project the possibilities to improve feed efficiency at the genetic level are to be researched, discussed and if possible evaluated.


### Validation of Predicted Breeding Values
Legarra and Reverter (2018) state that the introduction of new methods for genetic and genomic evaluations raise the question on how to assess and how to validate the used models in terms of their quality. This increases the requirement for tools to do these model assessments and validations. Even more so since with the implementation of genomic selection, riskier decisions such as selecting young and unproven bulls are encouraged. 

Reverter et al. (1991) have shown that the amount of change in EBV from one evaluation to the next one can be predicted. This predictability can be used to compute how likely a given change could be. Based on such computations, validation checks for the routine evaluation can be derived.

Simple validation statistics based on the LR-method (Legarra and Reverter (2018)) are already available in the R-package qzwslrm (Validation of Predicted Breeding Values Using the LR-Method ). The available validation statistics can be extended and can specifically be tuned towards a comparison between pedigree-based traditional BLUP breeding values and genomic BLUP breeding values (GBLUP) 


### Index to Rank Beef Bulls for their Use on Dairy Cows
Swiss AI Industry has expressed a long standing interest for a tool that provides a solid ranking of beef bulls for use on dairy cows. The reason for this requirement is the strong interest of dairy farmers in beef bulls (Beef on Dairy)

Berry et al (2019) have proposed such an index for the dairy and beef industry in Ireland. 

The approach by Berry et al (2019) would need to be adapted to the Swiss context

After a literature search on the topic of using beef bulls on dairy cows from the same authors, a ranking index for the Swiss context is to be proposed


---

If you are interested in any of the following topics or if you have questions, please contact Peter von Rohr (Qualitas AG) Chamerstrasse 56, 6300 Zug; peter.vonrohr@qualitasag.ch, Tel 041 768 92 21.


```{r echo=FALSE, results='asis'}
cat(rmdhelp::get_latest_change())
```

