---
title: "R Notebook"
output: html_notebook
---

## Data

```{r, message=FALSE, warning=FALSE}
s_ex03p03_data_path <- "https://charlotte-ngs.github.io/asmss2022/data/asm_bw_flem.csv"
s_ex03p03_data_path <- file.path(here::here(), "docs", "data", "asm_bw_flem.csv")
tbl_ex03p03_data <- readr::read_csv(file = s_ex03p03_data_path)
```


## Least Squares Normal Equations
In this section, we are computing one solution for the Least Squares Normal Equation. First sort the data according to the breed

```{r}
tbl_ex03p03_data <- tbl_ex03p03_data[order(tbl_ex03p03_data$Breed),]
```


```{r}
mat_X <- model.matrix(lm(`Body Weight` ~ 0 + Breed, data = tbl_ex03p03_data))
mat_X <- cbind(matrix(1, nrow = nrow(tbl_ex03p03_data), ncol = 1), mat_X)
dimnames(mat_X) <- NULL
mat_xtx <- crossprod(mat_X)
mat_xtx
```

The MP generalized inverse is obtained by 

```{r}
mat_xtx_ginv <- MASS::ginv(mat_xtx)
mat_xtx_ginv
```

The right hand side

```{r}
vec_y <- tbl_ex03p03_data$`Body Weight`
mat_xty <- crossprod(mat_X, vec_y)
mat_xty
```

One solution is obtained by

```{r}
mat_b0 <- crossprod(mat_xtx_ginv, mat_xty)
mat_b0
```

Setting the components of this solution vector $b^0$ equal to 

```{r, echo=FALSE, results='asis'}
vec_b0 <- c("\\mu", "\\alpha_1", "\\alpha_2", "\\alpha_3")
cat(paste0(rmdhelp::bcolumn_vector(pvec = vec_b0, ps_name = "b^0", ps_env = "$$"), collapse = ""))
```


## Contrasts

These results are consistent with the results obtained by `lm()`. If we wanted to see what type of estimable functions are used by the `lm()` function, it can be derived from the ouput of the function `contrasts()`. For our example with the Breed, we obtain the contrast matrix by 

```{r}
contrasts(as.factor(tbl_ex03p03_data$Breed))
```

This means that the effect reported as `Limousin` by `lm()` corresponds to the first estimable function and the effect reported as `Simmental` corresponds to the second estimable function. What the estimable functions are, we obtain from the computations that follow below. The effect for `Angus` is contained in the intercept which is just set to the mean body weight of all Angus animals. 

Adding to the above contrast matrix a row of all ones to the left side and inverting the extended matrix gives

```{r}
mat_contr <- contrasts(as.factor(tbl_ex03p03_data$Breed))
mat_contr <- cbind(matrix(1, nrow = nrow(mat_contr), ncol = 1), mat_contr)
mat_contr
```

Inverting the above shown contrast matrix yields the matrix with estimable functions

```{r}
mat_estf <- solve(mat_contr)
row.names(mat_estf)[1] <- "(Intercept)"
mat_estf
```

The frist row of the above matrix shows that the intercept corresponds to the mean of all Angus animals. This can be verified by

```{r}
n_inter_dummy <- mean(tbl_ex03p03_data[tbl_ex03p03_data$Breed == "Angus", ]$`Body Weight`)
n_inter_dummy
```

The linear model yields

```{r}
lm_ex03p03 <- lm(`Body Weight` ~ Breed, data = tbl_ex03p03_data)
coefficients(lm_ex03p03)["(Intercept)"]
```

The second and the third row of the above shown matrix of estimable functions contain the vector $q^T$ for the effect that are given as rownames. For example for the effect `Limousin` the row-vector $q^T$ is

```{r, echo=FALSE, results='asis'}
vec_q_LI <- mat_estf[2,]
cat(paste(rmdhelp::brow_vector(pvec = vec_q_LI, ps_name = "q_{LI}", ps_env = "$$")))
```

This means that the effect for Limousin corresponds to the difference between $\alpha_2 - \alpha_1$ which corresponds to 

```{r}
eff_LI <- mat_b0[3] - mat_b0[2]
eff_LI
```

The solution from `lm()` is

```{r}
coefficients(lm_ex03p03)["BreedLimousin"]
```

Similarly for the effect of Simmental. The estimable function is determined by the third row of the matrix of estimable functions. 

```{r, echo=FALSE, results='asis'}
vec_q_SI <- mat_estf[3,]
cat(paste(rmdhelp::brow_vector(pvec = vec_q_SI, ps_name = "q_{SI}", ps_env = "$$")))
```

That means, the effect for Simmental corresponds to the difference between $\alpha_3$ and $\alpha_1$. 

```{r}
eff_SI <- mat_b0[4] - mat_b0[2]
eff_SI
```

The solution from `lm()`

```{r}
coefficients(lm_ex03p03)["BreedSimmental"]
```


## Example from Video https://youtu.be/eVSIDf5w1_I

```{r}
data("OrchardSprays")
str(OrchardSprays)
```

Plot

```{r}
plot(decrease ~ treatment, data = OrchardSprays)
```

```{r}
model1 <- lm(decrease ~treatment, data = OrchardSprays)
summary(model1)
```

Use heterogeneous variance for different treatement classes using gls in model3
```{r}
library(nlme)
model2 <- gls(decrease~treatment, data = OrchardSprays)
model3 <- update(model2, weights=varIdent(form=~1|treatment))
AIC(model2, model3)
```

```{r}
summary(model3)
```

Succesive difference contrasts

Save old options
```{r}
opts <- options()
getOption("contrasts")
```

Change contrasts and run model

```{r}
library(MASS)
options(contrasts = c("contr.sdif", "contr.sdif"))
model4 <- gls(decrease ~ treatment, data = OrchardSprays)
model5 <- update(model4,  weights=varIdent(form=~1|treatment))
summary(model5)
```

Re-set options

```{r}
options(opts)
```

## Example from https://bookdown.org/pingapang9/linear_models_bookdown/chap-contrasts.html

```{r}
library(dplyr)
library(broom)
data("PlantGrowth")
PlantGrowth %>% 
  filter(group != "ctrl") %>% 
  lm(weight ~ group, .) %>%
  tidy()
```


## Example from https://marissabarlaz.github.io/portfolio/contrastcoding/
Examples from the mentioned website are applied to our extended dataset assigned to `tbl_ex03p03_data`. 

### Dummy Coding
The coding scheme that is most familiar is the dummy coding or the treatment coding. In that coding all levels are compares to a reference level. The reference is the level that is first in an alphabetical order. The contrast matrix for a factor with four levels looks as follows

```{r}
contr.treatment(4)
```

Checking the contrasts for the dummy variables `Breed` and `BCS` after conversion to factors, gives

```{r}
tbl_ex03p03_data$Breed <- as.factor(tbl_ex03p03_data$Breed)
tbl_ex03p03_data$BCS <- as.integer(tbl_ex03p03_data$BCS)
tbl_ex03p03_data$BCS <- as.factor(tbl_ex03p03_data$BCS)
```

Contrasts for `Breed`

```{r}
contrasts(tbl_ex03p03_data$Breed)
```

Extending the contrasts matrix

```{r}
con_mat <- contrasts(tbl_ex03p03_data$Breed)
con_mat
```

The extension is by a row of ones which stands for the intercept

```{r}
con_mat <- cbind(matrix(1, nrow = nrow(con_mat), ncol = 1), con_mat)
con_mat
```

This is a symmetric matrix. What is the inverse of this matrix

```{r}
est_mat <- solve(con_mat)
est_mat
```



Contrasts for `BCS` 

```{r}
contrasts(tbl_ex03p03_data$BCS)
```

The contrasts can be changed by an assignment

```{r}
nr_breed <- nlevels(tbl_ex03p03_data$Breed)
contrasts(tbl_ex03p03_data$Breed) <- contr.treatment(nr_breed)
lm_bwbrbcs <- lm(`Body Weight` ~ Breed + BCS, data = tbl_ex03p03_data)
summary(lm_bwbrbcs)
```

Changing to a different reference level

```{r}
contrasts(tbl_ex03p03_data$Breed) <- contr.treatment(nr_breed, base = nr_breed)
contrasts(tbl_ex03p03_data$Breed)
```

```{r}
lm_bwbrbcs2 <- lm(`Body Weight` ~ Breed + BCS, data = tbl_ex03p03_data)
summary(lm_bwbrbcs2)
```

## Consecutive Differences

```{r}
contr.treatment(3)
```

```{r}
contr.sdif(3)
```

Extending the sdif contrast

```{r}
mat_sdif <- contr.sdif(3)
mat_sdif <- cbind(matrix(1, nrow = nrow(mat_sdif), ncol = 1), mat_sdif)
mat_sdif
```

Estimable function

```{r}
mat_est_sdif <- solve(mat_sdif)
mat_est_sdif
```

```{r}
contr.sum(3)
```

```{r}
mat_sum <- contr.sum(3)
mat_sum <- cbind(matrix(1, nrow = nrow(mat_sum), ncol = 1), mat_sum)
mat_sum
```

```{r}
mat_est_sum <- solve(mat_sum)
mat_est_sum
```




## Example from https://stats.oarc.ucla.edu/r/library/r-library-contrast-coding-systems-for-categorical-variables/

Reading the data

```{r}
hsb2 = read.table('https://stats.idre.ucla.edu/stat/data/hsb2.csv', header=T, sep=",")
str(hsb2)
```

Create a new variable

```{r}
hsb2$race.f = factor(hsb2$race, labels=c("Hispanic", "Asian", "African-Am", "Caucasian"))
```

Descriptive statistics

```{r}
tapply(hsb2$write, hsb2$race.f, mean)
```

Plot

```{r}
plot(hsb2$race.f, hsb2$write)
```

### Dummy Coding
Comparison of each level of categorical variable to a reference.

```{r}
(contrasts(hsb2$race.f) = contr.treatment(4))
```

```{r}
summary(lm(write ~ race.f, hsb2))
```


## Resources

https://www.r-bloggers.com/2015/01/using-and-interpreting-different-contrasts-in-linear-models-in-r/
https://rcompanion.org/rcompanion/h_01.html
https://bookdown.org/pingapang9/linear_models_bookdown/chap-contrasts.html
https://marissabarlaz.github.io/portfolio/contrastcoding/
https://www.uvm.edu/~statdhtx/StatPages/R/ContrastsInR.html
https://cran.r-project.org/web/packages/multcomp/vignettes/multcomp-examples.pdf
https://stats.stackexchange.com/questions/78354/what-is-a-contrast-matrix
https://stats.oarc.ucla.edu/r/library/r-library-contrast-coding-systems-for-categorical-variables/
https://talklab.psy.gla.ac.uk/simgen/index.html

https://youtu.be/eVSIDf5w1_I
https://youtu.be/qQHNRaBbR90
https://youtu.be/fXDNBeY2qp0
