---
title: Applied Statistical Methods - `r tools::toTitleCase(params$doctype)` 8
author: Peter von Rohr
date: 2022-04-25
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
    keep_md: false
    keep_tex: false
header-includes:
 \usepackage{longtable}
 \usepackage{float}
params:
  doctype:
    label: 'Document Type'
    value: 'solution'
    choices: ['exercise', 'solution', 'notebook']
  isonline:
    label: 'Online (y/n)'
    value: FALSE
    choices: [TRUE, FALSE]
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Problem 1: Repeated Measurements Data
```{r asm-ex08-p01-init, echo=FALSE}
if (params$isonline){
  s_asm_ex08_p01_data_path <- "https://charlotte-ngs.github.io/asmss2022/data/asm_bw_flem.csv"
} else {
  s_asm_ex08_p01_data_path <- file.path(here::here(), "docs", "data", "asm_bw_flem.csv")
}
n_nr_rep <- 5
n_sd_prop_bw <- 0.5
```

Simulate a dataset with repeated measurements of `Body Weight` and `Breed`. The following dataset can be used as a basis: 

```{r asm_ex08-p01-data-path-show, echo=FALSE}
cat(s_asm_ex08_p01_data_path, "\n")
```

The generated dataset should have the following properties

* For every observation, the ID of the animal, its `Body Weight` and its `Breed` should be contained in the dataset.
* Each animal of the given basis dataset should have `r n_nr_rep` repeated observations of `Body Weight` and `Breed`.
* The phenotypic variance of `Body Weight` within the repeated observations of one animal should be `r 100 * n_sd_prop_bw`% of the total phenotypic variance of `Body Weight` determined from the given basis dataset.


### Your Tasks

* Analyse the generated dataset with an ANOVA
* Try to see whether you can re-cover the used input data in the results of the analysis


<!-- your-solution-start
### Your Solution

* Read the given basis dataset
* Loop over the records in the basis dataset and add the required number of records
* Analyse the generated dataset with an ANOVA
* Assess the results and compare them with the input used in the simulation

---  your-solution-end -->


<!-- master-solution-start -->

### Solution

* Read the given basis dataset. First assign the variable with the datafile name

```{r, eval=params$isonline}
s_asm_ex08_p01_data_path <- "https://charlotte-ngs.github.io/asmss2022/data/asm_bw_flem.csv"
```

Read the data

```{r}
tbl_ex08_p01 <- readr::read_csv(file = s_asm_ex08_p01_data_path)
tbl_ex08_p01 <- dplyr::select(tbl_ex08_p01, Animal, `Body Weight`, Breed)
head(tbl_ex08_p01)
```


* Loop over the records in the basis dataset and add the required number of records

```{r}
set.seed(9875)
sd_bw <- sd(tbl_ex08_p01$`Body Weight`)
tbl_rep_obs_result <- NULL
for (idx in 1:nrow(tbl_ex08_p01)){
  tbl_rep_cur <- dplyr::bind_rows(tbl_ex08_p01[idx,],
                                  tibble::tibble(Animal = c(rep(tbl_ex08_p01$Animal[idx], n_nr_rep - 1)),
                                                 `Body Weight` = rnorm((n_nr_rep-1), 
                                                                       mean = tbl_ex08_p01$`Body Weight`[idx], 
                                                                       sd = n_sd_prop_bw * sd_bw),
                                                 Breed = c(rep(tbl_ex08_p01$Breed[idx], n_nr_rep-1))))
  if (is.null(tbl_rep_obs_result)){
    tbl_rep_obs_result <- tbl_rep_cur
  } else {
    tbl_rep_obs_result <- dplyr::bind_rows(tbl_rep_obs_result, tbl_rep_cur)
  }
}
head(tbl_rep_obs_result)
```

The generated dataset is written to a file, such that it will be available for Problem 2

```{r}
s_ex08_p01_rep_obs_data_dir <- file.path(here::here(), "docs", "data")
if (!dir.exists(s_ex08_p01_rep_obs_data_dir)) 
  dir.create(path = s_ex08_p01_rep_obs_data_dir, recursive = TRUE)
s_ex08_p01_rep_obs_data_path <- file.path(s_ex08_p01_rep_obs_data_dir, 
                                          "asm_ex08_p01_rep_obs.csv")
if (!file.exists(s_ex08_p01_rep_obs_data_path)) 
  readr::write_csv(tbl_rep_obs_result, file = s_ex08_p01_rep_obs_data_path)
```


* Analyse the generated dataset with an ANOVA

```{r}
tbl_rep_obs_result$Animal <- as.factor(tbl_rep_obs_result$Animal)
aov_ex08_p01 <- aov(`Body Weight` ~ Breed + Error(Animal), data = tbl_rep_obs_result)
summary(aov_ex08_p01)
```

* Assess the results and compare them with the input used in the simulation

The variance of the residuals in the original basis dataset is obtained from 

```{r}
lm_bw_breed <- lm(`Body Weight` ~ Breed, data = tbl_ex08_p01)
(smry_bw_breed <- summary(lm_bw_breed))
```

The estimate of the residual variance is `r smry_bw_breed$sigma` which is comparable to the value found by `aov()`. The used value for the variance within observations is given by 

```{r}
n_obs_var_bw <- var(tbl_ex08_p01$`Body Weight`)
n_obs_var_bw * n_sd_prop_bw^2
```

This variance is used for each of the animals in the base dataset. Hence in total the variance between observations has to be summed over all animal groups.

<!-- master-solution-end -->


## Problem 2: Random Effects Model
```{r ex08-p02-init, echo=FALSE}
s_ex08_p02_data_path <- "https://charlotte-ngs.github.io/asmss2022/data/asm_ex08_p01_rep_obs.csv"
if (!params$isonline)
  s_ex08_p02_data_path <- file.path(here::here(), "docs", "data", "asm_ex08_p01_rep_obs.csv")
```

Analyse the dataset generated in Problem 1 with a random effects model using the package `lme4`. If you had difficulties to solve Problem 1, then you can also use the following dataset.

```{r, echo=FALSE}
cat(s_ex08_p02_data_path, "\n")
```


<!-- your-solution-start
### Your Solution

* Read generated dataset from Problem 1
* Analyse the data using `lme4::lmer()`

---  your-solution-end -->


<!-- master-solution-start -->

### Solution

* Read generated dataset from Problem 1

```{r}
# read the data
tbl_ex08_p02 <- readr::read_csv(file = s_ex08_p02_data_path)
# convert animal and breed to factors
tbl_ex08_p02$Animal <- as.factor(tbl_ex08_p02$Animal)
tbl_ex08_p02$Breed <- as.factor(tbl_ex08_p02$Breed)
tbl_ex08_p02
```


* Analyse the data using `lme4::lmer()`

The mixed model analysis is done with 

```{r}
lmer_ex08_p02 <- lme4::lmer(`Body Weight` ~ Breed + (1|Animal), data = tbl_ex08_p02)
summary(lmer_ex08_p02)
```

Plots

```{r}
plot(lmer_ex08_p02)
```

<!-- master-solution-end -->


