---
title: Pedigree BLUP - Sire Model
author: Peter von Rohr
date: "`r Sys.Date()`"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Independence Assumption

So far

* random residuals: $var(e) = I * \sigma_e^2$ and
* random sire effects: $var(s) = I * \sigma_s^2$ 

In real livestock datasets, this is not realistic, because sires are related


## Example Dataset

```{r, echo=FALSE}
tbl_sire_model <- tibble::tibble(Animal = c(4:8),
                                 Sire   = c(1,3,1,4,3),
                                 Sex    = c("M","F","F","M","M"),
                                 WWG    = c(4.5, 2.9, 3.9, 3.5, 5.0))
knitr::kable(tbl_sire_model,
             booktabs = TRUE,
             longtable = TRUE)

```


## Relationship

* For son $i$ and sire $k$ of $i$

\begin{equation}
cov(s_i, s_k) = 1/2 * \sigma_s^2 \notag
\end{equation}


## Sire Relationship Matrix

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(pedigreemm)
ped_sire <- pedigree(sire = c(rep(NA,2), 1), 
                     dam = rep(NA,3), 
                     label = as.character(c(1,3,4)))
mat_A <- getA(ped = ped_sire)
```

```{r, echo=FALSE, results='asis'}
cat(paste0(rmdhelp::bmatrix(pmat = as.matrix(mat_A), ps_name = "A", ps_env = "$$"), collapse = "\n"), "\n")
```


## Sire Model

* `pedigreemm` 

```{r, echo=TRUE}
lmem_sire <- pedigreemm(
  formula = WWG ~ Sex + (1 | Sire), 
  data = tbl_sire_model,
  pedigree = list(Sire = ped_sire)
)
summary(lmem_sire)
```


## Mixed model equations

* `pedigreemm`: Cannot specify assumed variance components

\begin{equation}
\left[ 
\begin{array}{cc}
X^TX  &  X^TZ \\
Z^TX  &  Z^TZ + \lambda A_s^{-1}
\end{array}
\right]
\left[ 
\begin{array}{c}
\hat{b} \\
\hat{s}
\end{array}
\right]
= 
\left[ 
\begin{array}{c}
X^Ty \\
Z^Ty
\end{array}
\right]
\notag
\end{equation}
