---
title: Applied Statistical Methods - Notebook 8
author: Peter von Rohr
date: '2022-04-25'
output: html_notebook
params:
  doctype:
    label: Document Type
    value: solution
    choices:
    - exercise
    - solution
    - notebook
  isonline:
    label: Online (y/n)
    value: true
    choices:
    - true
    - false
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Problem 1: Repeated Measurements Data
```{r asm-ex08-p01-init, echo=FALSE}
if (params$isonline){
  s_asm_ex08_p01_data_path <- "https://charlotte-ngs.github.io/asmss2022/data/asm_bw_flem.csv"
} else {
  s_asm_ex08_p01_data_path <- file.path(here::here(), "docs", "data", "asm_bw_flem.csv")
}
n_nr_rep <- 5
n_sd_prop_bw <- 0.5
```

Simulate a dataset with repeated measurements of `Body Weight` and `Breed`. The following dataset can be used as a basis: 

```{r asm_ex08-p01-data-path-show, echo=FALSE}
cat(s_asm_ex08_p01_data_path, "\n")
```

The generated dataset should have the following properties

* For every observation, the ID of the animal, its `Body Weight` and its `Breed` should be contained in the dataset.
* Each animal of the given basis dataset should have `r n_nr_rep` repeated observations of `Body Weight` and `Breed`.
* The phenotypic variance of `Body Weight` within the repeated observations of one animal should be `r 100 * n_sd_prop_bw`% of the total phenotypic variance of `Body Weight` determined from the given basis dataset.


### Your Tasks

* Analyse the generated dataset with an ANOVA
* Try to see whether you can re-cover the used input data in the results of the analysis


### Your Solution

* Read the given basis dataset
* Loop over the records in the basis dataset and add the required number of records
* Analyse the generated dataset with an ANOVA
* Assess the results and compare them with the input used in the simulation





## Problem 2: Random Effects Model
```{r ex08-p02-init, echo=FALSE}
s_ex08_p02_data_path <- "https://charlotte-ngs.github.io/asmss2022/data/asm_ex08_p01_rep_obs.csv"
if (!params$isonline)
  s_ex08_p02_data_path <- file.path(here::here(), "docs", "data", "asm_ex08_p01_rep_obs.csv")
```

Analyse the dataset generated in Problem 1 with a random effects model using the package `lme4`. If you had difficulties to solve Problem 1, then you can also use the following dataset.

```{r, echo=FALSE}
cat(s_ex08_p02_data_path, "\n")
```


### Your Solution

* Read generated dataset from Problem 1
* Analyse the data using `lme4::lmer()`







```{r, echo=FALSE, results='asis'}
cat('\n---\n\n _Latest Changes: ', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), ' (', Sys.info()['user'], ')_\n', sep = '')
```
 
