---
title: "Generate Data for Problem 1 of asm_ex07"
output: html_notebook
params:
  isonline:
    label: 'Online (y/n)'
    value: FALSE
    choices: [TRUE, FALSE]
---

The data for Problem 1 of asm_ex07 is a simulated dataset that uses results of the regression of body weight on breast circumference as input. 


## Regression Analysis
Start with a regression analysis on the small dataset.

```{r sim-data-ex07-p01}
# first do a regression analysis on the small dataset with 10 records
if (params$isonline){
  s_ex05p02_data_path <- "https://charlotte-ngs.github.io/asmss2022/data/asm_bw_flem.csv"
} else {
  s_ex05p02_data_path <- file.path(here::here(), "docs", "data", "asm_bw_flem.csv")
}
tbl_bwbc <- readr::read_csv(file = s_ex05p02_data_path)
# extract input for simulation from results of regression analysis
lm_bwbc <- lm(`Body Weight` ~ `Breast Circumference`, data = tbl_bwbc)
```


## Input for Simulation
Extract input for simulation from results of regression analysis

```{r}
b0 <- coefficients(lm_bwbc)["(Intercept)"]
b1 <- coefficients(lm_bwbc)["`Breast Circumference`"]
mean_bc <- mean(tbl_bwbc$`Breast Circumference`)
sd_bc <- sd(tbl_bwbc$`Breast Circumference`)
sry_bwbc <- summary(lm_bwbc)
sd_res <- sry_bwbc$sigma
```


## Simulation Function
The following function is used for simulating data

```{r}
simulate_bwbc <- function(pn_nrobs, pn_b0, pn_b1, pn_mean_x, pn_sd_x, pn_sd_res){
  vec_bc <- rnorm(pn_nrobs, mean = pn_mean_x, sd = pn_sd_x)
  vec_bw <- pn_b0 + pn_b1 * vec_bc + rnorm(pn_nrobs, mean = 0, sd = pn_sd_res)
  tbl_result <- tibble::tibble(Animal = c(1:pn_nrobs),
                               BC = vec_bc,
                               BW = vec_bw)
  return(tbl_result)
}
```

The following chunk generates the dataset

```{r}
set.seed(1762)
n_nr_obs <- 30
tbl_ex07p01 <- simulate_bwbc(pn_nrobs = n_nr_obs, 
                     pn_b0 = b0, 
                     pn_b1 = b1, 
                     pn_mean_x = mean_bc, 
                     pn_sd_x = sd_bc, 
                     pn_sd_res = sd_res)
tbl_ex07p01
```

Add random variables

```{r}
# height
n_mean_hei <- 130
n_sd_hei <- 1.7
tbl_ex07p01$HEI <- rnorm(nrow(tbl_ex07p01), mean = n_mean_hei, sd = n_sd_hei)
# BCS
vec_bcs_level <- c(1:5)
tbl_ex07p01$BCS <- sample(vec_bcs_level, size = n_nr_obs, replace = TRUE)
```

Add correlated breed

```{r}
vec_breed <- c("Angus", "Simmental", "Limousin")
n_nr_ani_per_breed <- floor(n_nr_obs / length(vec_breed))
tbl_ex07p01 <- tbl_ex07p01[order(tbl_ex07p01$BW),]
# initialize breed factor
tbl_ex07p01$Breed <- rep(NA, n_nr_obs)
# distribute breed levels
for (i in seq_along(vec_breed)){
  tbl_ex07p01$Breed[((i-1)*n_nr_ani_per_breed+1):(i*n_nr_ani_per_breed)] <- vec_breed[i]  
}
# add latest breed to last records
if ((n_nr_ani_per_breed * length(vec_breed)) < n_nr_obs){
  tbl_ex07p01$Breed[(n_nr_ani_per_breed * length(vec_breed) + 1):n_nr_obs] <- vec_breed[length(vec_breed)]
}
```

Reorder dataset according to animals

```{r}
tbl_ex07p01 <- dplyr::arrange(tbl_ex07p01, Animal)
tbl_ex07p01
```


## Write Data to File
Write data to file, if file does not exist

```{r}
s_ex07p01_path <- file.path(here::here(), "docs", "data", "asm_bw_mod_sel.csv")
if (!file.exists(s_ex07p01_path))
  readr::write_csv(tbl_ex07p01, file = s_ex07p01_path)
```

