---
title: "Data Set on Fixed Linear Effect Models"
output: html_notebook
---

# Disclaimer
The dataset used for multiple regression is extended by breed and is later used in fixed linear effect models.


# Read Multiple Regression Dataset
The existing dataset for multiple regression is read

```{r}
s_data_dir <- file.path(here::here(), "docs", "data")
list.files(path = s_data_dir)
s_data_path <- file.path(s_data_dir, "asm_bw_mult_reg.csv")
tbl_mult_reg <- readr::read_csv(file = s_data_path)
tbl_mult_reg
```


# Breed of Animals
The breed of the animals is manually assigned

```{r}
tbl_mult_reg[order(tbl_mult_reg$`Body Weight`),]$Animal
```

The breed is assigned according to the ranking of body weights

```{r}
tbl_mult_reg$Breed[tbl_mult_reg[order(tbl_mult_reg$`Body Weight`),]$Animal] <- c(rep("Angus", 3), rep("Simmental", 3), rep("Limousin", 4))
```

The extended dataset

```{r}
tbl_mult_reg
```


# Write Extended Dataset
The extended dataset is written to a file

```{r}
s_flem_data_path <- file.path(s_data_dir, "asm_bw_flem.csv")
readr::write_csv(tbl_mult_reg, file = s_flem_data_path)

```

