---
title: Applied Statistical Methods - `r tools::toTitleCase(params$doctype)` 1
author: 
date: 2022-02-25
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
    keep_md: false
    keep_tex: false
header-includes:
 \usepackage{longtable}
 \usepackage{float}
params:
  doctype:
    label: 'Document Type'
    value: 'solution'
    choices: ['exercise', 'solution', 'notebook']
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Problem 1: Regression Analysis
The following dataset on body weight and on futher observations on a number of animals is given.

```{r tbl-ex01-p01-data, echo=FALSE, message=FALSE, warning=FALSE}
s_ex01_p01_path <- "https://charlotte-ngs.github.io/asmss2022/data/asm_bw_mult_reg.csv"
tbl_ex01_p01_data <- readr::read_csv(file = s_ex01_p01_path)
knitr::kable(tbl_ex01_p01_data,
             booktabs = TRUE,
             longtable = TRUE,
             escape = FALSE)
```

The data can be read from `r s_ex01_p01_path`

### Tasks

* Build a regression model of body weight on the other observations using the dataset given above.
* Set up the matrix $\mathbf{X}$ and the vectors $\mathbf{y}$, $\mathbf{b}$ and $\mathbf{e}$.
* Compute estimate for the regression coefficients in the model defined above.


<!-- your-solution-start
### Your Solution

---  your-solution-end -->


<!-- master-solution-start -->

### Solution

First, the data has to be read

```{r, message=FALSE, warning=FALSE}
s_sol01_p01_path <- "https://charlotte-ngs.github.io/asmss2022/data/asm_bw_mult_reg.csv"
tbl_sol01_p01 <- readr::read_csv(file = s_sol01_p01_path)
```

The regression model can be stated as 

$$ \mathbf{y} = \mathbf{X}\mathbf{b} + \mathbf{e}$$

where 

```{r, echo=FALSE, results='asis'}
n_nr_obs <- nrow(tbl_sol01_p01)
n_nr_pred <- 4
vec_y <- tbl_sol01_p01$`Body Weight`
mat_X <- matrix(c(rep(1, n_nr_obs), 
                  tbl_sol01_p01$`Breast Circumference`, 
                  tbl_sol01_p01$BCS, 
                  tbl_sol01_p01$HEI), ncol = n_nr_pred)
vec_b <- rmdhelp::vecGetVecElem(psBaseElement = "b", pnVecLen = n_nr_pred, psResult = "latex")
vec_e <- rmdhelp::vecGetVecElem(psBaseElement = "e", pnVecLen = n_nr_obs, psResult = "latex")
cat("$$\n")
cat(paste0(rmdhelp::bmatrix(pmat = mat_X, ps_name = "X"), collapse = ""), "\n")
cat("\\text{, }\n")
cat(paste0(rmdhelp::bcolumn_vector(pvec = vec_y, ps_name = "y"), collapse = ""), "\n")
cat("\\text{, }\n")
cat(paste0(rmdhelp::bcolumn_vector(pvec=vec_e, ps_name = "e"), collapse = ""), "\n")
cat("\\text{, }\n")
cat(paste0(rmdhelp::bcolumn_vector(pvec = vec_b, ps_name = "b"), collapse = ""), "\n")
cat("$$\n")
```

The solution for the estimates $\hat{b}$ can be computed as

$$\hat{b} = (X^TX)X^Ty $$

```{r}
mat_xtx <- crossprod(mat_X)
mat_xty <- crossprod(mat_X, vec_y)
mat_sol_bhat <- solve(mat_xtx, mat_xty)
```

The solution vector is

```{r, echo=FALSE, results='asis'}
cat(paste0(rmdhelp::bmatrix(pmat = round(mat_sol_bhat, digits = 3), ps_name = "\\hat{b}", ps_env = "$$"), collapse = ""), "\n")
```


<!-- master-solution-end -->


## Problem 2

<!-- your-solution-start
### Your Solution

---  your-solution-end -->


<!-- master-solution-start -->

### Solution

<!-- master-solution-end -->
 
