---
title: Applied Statistical Methods - `r tools::toTitleCase(params$doctype)` 9
author: Peter von Rohr
date: 2022-05-02
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
    keep_md: false
    keep_tex: false
header-includes:
 \usepackage{longtable}
 \usepackage{float}
params:
  doctype:
    label: 'Document Type'
    value: 'solution'
    choices: ['exercise', 'solution', 'notebook']
  isonline:
    label: 'Online (y/n)'
    value: TRUE
    choices: [TRUE, FALSE]
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Problem 1: Milk Dataset
Use the dataset `milk` from package `pedigreemm` and fit a sire model to each of the response variables (`milk`, `fat`, `prot` and `scs`) in the data. The dataset can be loaded using the command `pedigreemm::milk`. The other variables like `lact` and `herd` can be used as fixed effects. The `sire` column is used as a random effect. For this analysis, we assume that sires are unrelated. 

### Your Tasks
* Analyse the `milk` dataset from package `pedigreemm` using the function `lme4::lmer()` for all given response variables. You can use the same model for each of the responses.
* Compute the estimated heritability for each response variable, using the fact that the heritability $h^2$ can be computed from the variance $\sigma_s^2$ of the sire effects and the phenotypic variance $\sigma_p^2$ with the formula 

$$h^2 = \frac{4* \sigma_s^2}{\sigma_p^2} $$

* Compute the summary statistic using the function `summary()` of all the predicted sire breeding values. Solutions for the sire breeding values are obtained using the function `ranef()`

<!-- your-solution-start
### Your Solution

* Assign milk dataset to a tibble
* Analyse the data for each of the responses
* Compute estimated heritability
* Obtain summary statistics for predicted sire breeding values

---  your-solution-end -->


<!-- master-solution-start -->

### Solution

* Assign milk dataset to a tibble

```{r}
tbl_milk <- tibble::as_tibble(pedigreemm::milk)
```

* Analyse the data for each of the responses. The responses are pre-defined and must appear as column names

```{r}
vec_resp_milk <- c("milk", "fat", "prot", "scs")
# check
if (!all(is.element(vec_resp_milk, colnames(tbl_milk)))){
  stop(" *** ERROR: Response not in dataset: ", 
       vec_resp_milk[!is.element(vec_resp_milk, colnames(tbl_milk))])
}
```

For each of the responses run an analysis

```{r}
vec_fix_fact_milk <- c("lact", "herd")
vec_fix_cov_milk <- c("dim")
s_rand_eff_milk <- "sire"
s_rhs <- paste0(c(paste0(vec_fix_fact_milk, collapse = " + "),
              paste0(vec_fix_cov_milk, collapse = " + "),
              paste0("(1|", s_rand_eff_milk, ")", collapse = "")),
              collapse = " + ")

```

Convert fixed effects to factors

```{r}
# convert fixed effects to factors
for (f in vec_fix_fact_milk){
  tbl_milk[[f]] <- as.factor(tbl_milk[[f]])
}
```

Construct a function that runs an analysis for a given response

```{r}
run_lmer_single_response <- function(ps_resp, ps_rhs, ptbl_data){
  # check
  # formula
  s_formula <- paste(ps_resp, ps_rhs, sep = " ~ ")
  frma_lmem <- as.formula(s_formula)
  return(lme4::lmer(frma_lmem, data = ptbl_data))
}
```

Let the analysis be done for all responses

```{r}
l_lmer_results <- lapply(vec_resp_milk, 
                         run_lmer_single_response, 
                         ps_rhs = s_rhs, 
                         ptbl_data = tbl_milk)
```

Print all the summaries as results

```{r}
for (l in l_lmer_results){
  print(summary(l))
}
```


* Compute estimated heritability. For the sire model, the heritability $h^2$ is defined as 

$$h^2 = \frac{4*\sigma_s^2}{\sigma_s^2 + \sigma_e^2}$$
The estimates for the sire variance $\sigma_s^2$ and $\sigma_e^2$ can be extracted from the summary of the `lmer` result.

```{r}
for (l in l_lmer_results){
  smry_l <- summary(l)
  # getting sire std. dev
  n_sire_var <- attr(smry_l$varcor[[1]], "stddev")[[1]]
  n_res_var <- smry_l$sigma
  n_h2 <- 4*n_sire_var / (n_sire_var+n_res_var)
  s_trait <- unlist(strsplit(as.character(smry_l$call)[2], split = " ~ ", fixed = TRUE))[1]
  cat("Heritability for trait: ", s_trait, ": ", n_h2, "\n")
}
```


* Obtain summary statistics for predicted sire breeding values. The breeding values can be obtained by the function `raneff()`

```{r}
for (l in l_lmer_results){
  smry_l <- summary(l)
  s_trait <- unlist(strsplit(as.character(smry_l$call)[2], split = " ~ ", fixed = TRUE))[1]
  cat("Summary statistics of breeding values for trait: ", s_trait, "\n")
  print(summary(lme4::ranef(l)$sire[[1]]))
}
```



<!-- master-solution-end -->


