---
title: Applied Statistical Methods - Notebook 4
author: Peter von Rohr
date: 2022-03-16
output: html_notebook
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
s_ex04p01_data_path <- "https://charlotte-ngs.github.io/asmss2022/data/asm_bw_flem.csv"
```


## Problem 1: Overfitting
Use the extended dataset on `Body Weight` of animals and fit all the variables and the factor breed. Compare the result with a regression that uses only `Breast Circumference` or with the linear model that only uses the factor `Breed`. The data set is available from: `r s_ex04p01_data_path` 



### Your Solution

* Read the data

* Fit the full model

* Fit the model with only `Breast Circumference`

* Fit the model with only `Breed`






## Problem 2: Plotting
The first step before doing any analysis should always be to plot the data which helps to visualise the internal structure of a dataset. A very instructive plot is the so-called `pairs`-plot. This plot can be done using the function `pairs()`. The task of this problem is to create a `pairs`-plot for the extended dataset on `Body Weight` of animals. The input to the function `pairs()` must be all numeric. This means that the column containing the `Breed` in our dataset must be converted to a datatype called `factor`. This can be done using the function `as.factor()`. 

Results of linear models can also be plotted. In such plots, we are mainly interested in the behavior of the residuals. Hence, fit a linear regression model between `Body Weight` and `Breast Circumference` and plot the resulting linear model object.

### Your Solution

* Read the dataset

* Convert the breed column to a factor

* Create a `pairs`-plot

* Fit the linear regression model

* Plot the result



 


```{r, echo=FALSE, results='asis'}
cat('\n---\n\n _Latest Changes: ', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), ' (', Sys.info()['user'], ')_\n', sep = '')
```
 
