---
title: Applied Statistical Methods - `r tools::toTitleCase(params$doctype)` 8
author: Peter von Rohr
date: 2022-04-25
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
    keep_md: false
    keep_tex: false
header-includes:
 \usepackage{longtable}
 \usepackage{float}
params:
  doctype:
    label: 'Document Type'
    value: 'solution'
    choices: ['exercise', 'solution', 'notebook']
  isonline:
    label: 'Online (y/n)'
    value: TRUE
    choices: [TRUE, FALSE]
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Problem 1: Repeated Measurements Data
```{r asm-ex08-p01-init, echo=FALSE}
if (params$isonline){
  s_asm_ex08_p01_data_path <- "https://charlotte-ngs.github.io/asmss2022/data/asm_bw_flem.csv"
} else {
  s_asm_ex08_p01_data_path <- file.path(here::here(), "docs", "data", "asm_bw_flem.csv")
}
n_nr_rep <- 5
n_sd_prop_bw <- 0.5
```

Simulate a dataset with repeated measurements of `Body Weight` and `Breed`. The following dataset can be used as a basis: 

```{r asm_ex08-p01-data-path-show, echo=FALSE}
cat(s_asm_ex08_p01_data_path, "\n")
```

The generated dataset should have the following properties

* For every observation, the ID of the animal, its `Body Weight` and its `Breed` should be contained in the dataset.
* Each animal of the given basis dataset should have `r n_nr_rep` repeated observations of `Body Weight` and `Breed`.
* The phenotypic variance of `Body Weight` within the repeated observations of one animal should be `r 100 * n_sd_prop_bw`% of the total phenotypic variance of `Body Weight` determined from the given basis dataset.


### Your Tasks

* Analyse the generated dataset with an ANOVA
* Try to see whether you can re-cover the used input data in the results of the analysis


<!-- your-solution-start
### Your Solution

* Read the given basis dataset
* Loop over the records in the basis dataset and add the required number of records
* Analyse the generated dataset with an ANOVA
* Assess the results and compare them with the input used in the simulation

---  your-solution-end -->


<!-- master-solution-start -->

### Solution

<!-- master-solution-end -->


## Problem 2: Random Effects Model
Analyse the dataset generated in Problem 1 with a random effects model using the package `lme4`.


<!-- your-solution-start
### Your Solution

* Read generated dataset from Problem 1
* Analyse the data using `lme4::lmer()`

---  your-solution-end -->


<!-- master-solution-start -->

### Solution

<!-- master-solution-end -->


