---
title: Applied Statistical Methods - Notebook 9
author: Peter von Rohr
date: '2022-05-02'
output: html_notebook
params:
  doctype:
    label: Document Type
    value: solution
    choices:
    - exercise
    - solution
    - notebook
  isonline:
    label: Online (y/n)
    value: true
    choices:
    - true
    - false
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Problem 1: Milk Dataset
Use the dataset `milk` from package `pedigreemm` and fit a sire model to each of the response variables (`milk`, `fat`, `prot` and `scs`) in the data. The dataset can be loaded using the command `pedigreemm::milk`. The other variables like `lact` and `herd` can be used as fixed effects. The `sire` column is used as a random effect. For this analysis, we assume that sires are unrelated. 

### Your Tasks
* Analyse the `milk` dataset from package `pedigreemm` using the function `lme4::lmer()` for all given response variables. You can use the same model for each of the responses.
* Compute the estimated heritability for each response variable, using the fact that the heritability $h^2$ can be computed from the variance $\sigma_s^2$ of the sire effects and the phenotypic variance $\sigma_p^2$ with the formula 

$$h^2 = \frac{4* \sigma_s^2}{\sigma_p^2} $$

* Compute the summary statistic using the function `summary()` of all the predicted sire breeding values. Solutions for the sire breeding values are obtained using the function `ranef()`

### Your Solution

* Assign milk dataset to a tibble
* Analyse the data for each of the responses
* Compute estimated heritability
* Obtain summary statistics for predicted sire breeding values







```{r, echo=FALSE, results='asis'}
cat('\n---\n\n _Latest Changes: ', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), ' (', Sys.info()['user'], ')_\n', sep = '')
```
 
