---
title: ANOVA and Random Effects Models
author: Peter von Rohr
date: "2022-04-25"
output: beamer_presentation
params:
  isonline:
    label: 'Online (y/n)'
    value: FALSE
    choices: [TRUE, FALSE]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Repeated Measurements

* Same characteristics (`Body Weight`, `BC`, ...) measured multiple times for the same animal

```{r, echo=FALSE, warning=FALSE, message=FALSE}
if (params$isonline){
  s_rep_obs_path <- "https://charlotte-ngs.github.io/asmss2022/data/asm_bw_bc_rep_obs.csv"
} else {
  s_rep_obs_path <- file.path(here::here(), "docs", "data", "asm_bw_bc_rep_obs.csv")
}
tbl_rep_obs <- readr::read_csv(file = s_rep_obs_path)
knitr::kable(head(tbl_rep_obs, 8),
             booktabs = TRUE,
             longtable = TRUE)
```

* Column `Animal` no longer just a counter, it becomes a model factor

## Properties 

* What is the benefit of repeated measurements?
* Equivalent to more animals in the dataset?
* Assumption: repeated observations of the same animal are more "similar"
* No longer independence of observations


## Data Scatterplot

```{r, echo=FALSE, message=FALSE, warning=FALSE, out.width="100%"}
tbl_rep_obs$Animal <- as.factor(tbl_rep_obs$Animal)
ggplot2::ggplot(data = tbl_rep_obs, 
                mapping = ggplot2::aes(x = `Breast Circumference`,
                                       y = `Body Weight`)) + 
  ggplot2::geom_point(ggplot2::aes(color = Animal, 
                                   size = 3))

```


## Statistical Analysis

* Technically, same regression analysis is possible

```{r}
lm_rep_obs <- lm(`Body Weight` ~ `Breast Circumference`, data = tbl_rep_obs)
```

* But assumption of independence is violoted
* Consequence of independence

$$var(\mathbf{e}) = \mathbf{I} * \sigma_e^2$$
* Check residuals plot



## No Repeated Measures


