---
title: Applied Statistical Methods - `r tools::toTitleCase(params$doctype)` 9
author: Peter von Rohr
date: 2022-05-02
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
    keep_md: false
    keep_tex: false
header-includes:
 \usepackage{longtable}
 \usepackage{float}
params:
  doctype:
    label: 'Document Type'
    value: 'solution'
    choices: ['exercise', 'solution', 'notebook']
  isonline:
    label: 'Online (y/n)'
    value: TRUE
    choices: [TRUE, FALSE]
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Problem 1: Milk Dataset
Use the dataset `milk` from package `pedigreemm` and fit a sire model to each of the response variables (`milk`, `fat`, `prot` and `scs`) in the data. The dataset can be loaded using the command `pedigreemm::milk`. The other variables like `lact` and `herd` can be used as fixed effects. The `sire` column is used as a random effect. For this analysis, we assume that sires are unrelated. 

### Your Tasks
* Analyse the `milk` dataset from package `pedigreemm` using the function `lme4::lmer()` for all given response variables. You can use the same model for each of the responses.
* Compute the estimated heritability for each response variable, using the fact that the heritability $h^2$ can be computed from the variance $\sigma_s^2$ of the sire effects and the phenotypic variance $\sigma_p^2$ with the formula 

$$h^2 = \frac{4* \sigma_s^2}{\sigma_p^2} $$

* Compute the summary statistic using the function `summary()` of all the predicted sire breeding values. Solutions for the sire breeding values are obtained using the function `ranef()`

<!-- your-solution-start
### Your Solution

* Assign milk dataset to a tibble
* Analyse the data for each of the responses
* Compute estimated heritability
* Obtain summary statistics for predicted sire breeding values

---  your-solution-end -->


<!-- master-solution-start -->

### Solution

* Assign milk dataset to a tibble

```{r}
tbl_milk <- tibble::as.tibble(pedigreemm::milk)
```

* Analyse the data for each of the responses. The responses are pre-defined and must appear as column names

```{r}
vec_resp_milk <- c("milk", "fat", "prot", "scs")
# check
if (!all(is.element(vec_resp_milk, colnames(tbl_milk)))){
  stop(" *** ERROR: Response not in dataset: ", vec_resp_milk[!is.element(vec_resp_milk, colnames(tbl_milk))])
}
```

For each of the responses run an analysis




* Compute estimated heritability
* Obtain summary statistics for predicted sire breeding values


<!-- master-solution-end -->


